---
title: "Cases & Deaths data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite)
library(ggplot2)
library(readr)
library(outbreakinfo)
```

# Accessing the data in outbreak.info's Variant Tracker
Outbreak.info's Cases & Deaths Tracker allows you to compare trends in COVID-19 cases and deaths by location over time. We will walk you through how to run some essential queries. We can help you to answer questions like:

- [How do I retrieve location names and codes used in the epi functions?](#how-to-retrieve-location-names-and-codes)
- [How do I pull epidemiological data by location or according to my own specs?](#how-to-retrieve-epi-data-1)
- [How do I plot epidemiological data?](#how-to-plot-epi-data)

------

## Before we start...
Import the packages we'll use to access the data.

Note that unlike the [SARS-CoV-2 Variant Prevalence functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-sars-cov-variant-prevalence), none of the Cases & Deaths functions or anything in this vignette require GISAID authentication through the `authenticateUser()` function.
```{r pkgs, message = FALSE, results='hide'}
# Package imports
library(outbreakinfo)

```

The following is a data dictionary for your reference:
```{r, eval=T, echo=T}
knitr::kable(epidemiologyDataDictionary())
```

------

## How to retrieve location names and codes

#### **getISO3**
**Description:** 

Get ISO3 codes for World Bank regions, countries, states/provinces, metropolitan areas, and/or counties.

**Usage:**

```{r, eval=F, echo=T}
getISO3(locations_to_search)
```

**Arguments:**

locations_to_search: a `vector` or `list` of location (World Bank region, country, state/province, metropolitan area, county) name(s)

**Example:**

```{r}
iso3=getISO3(c("United States of America", "Brazil"))
print(iso3)
```

------

#### **searchLocations**
**Description:** 

Get exact spelling of locations at the same administrative level.

**Usage:**

```{r, eval=F, echo=T}
searchLocations(locations_to_search, admin_level)
```

**Arguments:**

locations_to_search: a `vector` or `list` of location (World Bank region, country, state/province, metropolitan area, county) name(s) at the same administrative level

admin_level: an `integer` representing an administrative level (World Bank regions = -1, countries = 0, states/provinces = 1, metropolitan areas = 1.5, counties = 2)

**Example:**

```{r}
names=searchLocations(c("California", "Florida"), admin_level=1)
print(names)
```

------

## How to retrieve epi data

#### **getEpiData**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) according to user specifications.

**Usage:**

```{r, eval=F, echo=T}
getEpiData(name=NULL, location_id=NULL, wb_region=NULL, country_name=NULL, state_name=NULL, admin_level=NULL, date=NULL, mostRecent=NULL, fields=NULL, sort, size=1000)
```

**Arguments:**

name: a `vector` or `list` of location (World Bank region, country, state/province, metropolitan area, county) name(s)

location_id: a `vector` or `list` of ISO3 codes representing locations (World Bank region, country, state/province, metropolitan area, county)

wb_region: a `vector` or `list` of World Bank region name(s)

country_name: a `vector` or `list` of country name(s)

state_name: a `vector` or `list` of state name(s)

admin_level: an `integer` representing an administrative level (World Bank regions = -1, countries = 0, states/provinces = 1, metropolitan areas = 1.5, counties = 2)

date: a `vector` or `list` of date(s) as string(s) in YYYY-MM-DD format

mostRecent: a `Boolean` (T/F) indicating if all data should be returned or only most recent

fields: a `vector` of API fields to include in results

sort: a `string` representing parameter to sort results by

size: an `integer` representing size

**Example:**

```{r}
df=getEpiData(name="United States of America", date="2020-07-01")
df[!duplicated(df$name), c("name", "date", "confirmed", "dead")]
```

------

#### **getLocationData**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for specified locaton(s) (World Bank region, country, state/province, metropolitan area, county).

**Usage:**

```{r, eval=F, echo=T}
getLocationData(location_names)
```

**Arguments:**

location_names: a `vector` or `list` of location (World Bank region, country, state/province, metropolitan area, county) name(s)

**Example:**

```{r}
df=getLocationData(location_names=c("Brazil", "San Diego"))
df[!duplicated(df$name), c("name", "date", "confirmed")]
```

--------

#### **getAdmn0**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all countries.

**Usage:**

```{r, eval=F, echo=T}
getAdmn0()
```

**Arguments:**

none

**Example:**

```{r, eval=F, echo=T}
world_df=getAdmn0()
length(unique(world_df$name))
```

--------

#### **getCountryByRegion**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all countries in one or more World Bank regions.

**Usage:**

```{r, eval=F, echo=T}
getCountryByRegion(wb_regions)
```

**Arguments:**

wb_regions:  a `vector` or `list` of World Bank region names

**Example:**

```{r}
region_df=getCountryByRegion("South Asia")
print(unique(region_df$name))
```

------

#### **getAdmn1ByCountry**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all states/provinces in one or more countries.

**Usage:**

```{r, eval=F, echo=T}
getAdmn1ByCountry(countries)
```

**Arguments:**

countries:  a `vector` or `list` of country names

**Example:**

```{r}
state_df=getAdmn1ByCountry("India")
print(unique(state_df$name))
```

--------

#### **getMetroByCountry**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all metropolitan areas in the United States of America.

**Usage:**

```{r, eval=F, echo=T}
getMetroByCountry()
```

**Arguments:**

none

**Example:**

```{r, eval=F, echo=T}
metro_df=getMetroByCountry()
```

--------

#### **getAdmn2ByCountry**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all counties in the United States of America.

**Usage:**

```{r, eval=F, echo=T}
getAdmn2ByCountry()
```

**Arguments:**

none

**Example:**

```{r, eval=F, echo=T}
county_df=getAdmn2ByCountry()
```

---------

#### **getAdmn2ByState**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all counties in given state(s).

**Usage:**

```{r, eval=F, echo=T}
getAdmn2ByState(states)
```

**Arguments:**

states: a `vector` or `list` of state names

**Example:**

```{r}
ca_df=getAdmn2ByState("California")
length(unique(ca_df$name))
```

--------

#### **getByAdmnLevel**
**Description:** 

Retrieve a `dataframe` of up-to-date COVID-19 data from [outbreak.info](https://outbreak.info/) for all locations at a specified administrative level.

**Usage:**

```{r, eval=F, echo=T}
getByAdmnLevel(admin_level)
```

**Arguments:**

admin_level: an `integer` representing an administrative level (World Bank regions = -1, countries = 0, states/provinces = 1, metropolitan areas = 1.5, counties = 2)

**Example:**

```{r}
admin_df=getByAdmnLevel(-1)
print(unique(admin_df$name))
```

--------

## How to plot epi data

#### **plotEpiData**
**Description:** 

Plot a metric of interest using up-to-date COVID-19 data using data from [outbreak.info](https://outbreak.info/) for location(s) of interest (World Bank region, country, state/province, metropolitan area, county)

**Usage:**

```{r, eval=F, echo=T}
plotEpiData(locations, variable)
```

**Arguments:**

location: a `vector` or `list` of location name(s)

variable: metric to plot

**Example:**

```{r}
p=plotEpiData(c("Brazil", "San Diego"), "confirmed_per_100k")
show(p)
```

------

