---
title: "Introduction to outbreakinfo"
# author: "Laura Hughes"
date: "`r format(Sys.Date(), "%d %B %Y")`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to outbreakinfo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
[outbreak.info](https://outbreak.info) is a project to enable the tracking of SARS-CoV-2 Variants within the COVID-19 pandemic. This R package offers access to the data we have gathered and calculated to replicate the visualizations on outbreak.info. Here, we'll outline some of the basic functionality of the R package to access the genomic (SARS-CoV-2 variant) data and epidemiology data (COVID-19 cases and deaths).

## SARS-CoV-2 variant prevalence

The core functionality within the suite of [Variant Prevalence Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-sars-cov-variant-prevalence) include accessing data for:

- [Variant prevalence over time](#prevalence-over-time)
- [Cumulative variant prevalence](#cumulative-variant-prevalence)
- [Calculating mutation prevalence within a variant](#muations-within-a-variant)

### Getting started
Before we start, we need to provide our GISAID credentials to access the genomic data.
If you don't have a GISAID account, you can [register](https://www.gisaid.org/registration/register/)
for one on their website. It may take a day or two for the account to become active.
```{r install, message = FALSE, results='hide', eval=FALSE}
# Install the package, if you haven't already, using devtools
devtools::install_github("outbreak-info/R-outbreak-info/outbreakinfo")
```

```{r pkgs, message = FALSE, results='hide'}
# Package imports
library(outbreakinfo)

# Not needed; just used to tidy some of the outputs
library(dplyr)
library(knitr)
```

```{r auth, message = FALSE, results='hide', eval=FALSE}
# Authenticate yourself using your GISAID credentials.
authenticateUser()
```

---

### Prevalence over time
Tracking how the prevalence of variants change over time is vital to understanding the evolution of SARS-CoV-2. We can calculate this prevalence for any lineage, mutation, combination of mutations, or lineage with added mutations. We'll access the data and then plot the results.

#### Lineage over time: B.1.1.7
```{r prev-time}
# Function to grab all the data for the prevalence of B.1.1.7 in San Diego
b117_san_diego = getPrevalence(pangolin_lineage = "B.1.1.7", location="San Diego")

# Accessing just the first row:
t(b117_san_diego[1,])

# Plotting it:
plotPrevalenceOverTime(b117_san_diego, title = "B.1.1.7 prevalence in San Diego")
```
Variables are described in the [Genomics Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/genomicsDataDictionary.html).

#### Customizing variant inputs
In addition to viewing the change in prevalence of a lineage over time, more complex variants can be queried. This includes **combinations of lineages**, like the [Delta variant](https://outbreak.info/situation-reports/delta), **lineages with additional mutation(s)**, like [B.1.1.7 with S:E484K](https://outbreak.info/situation-reports?muts=S%3AE484K&pango=B.1.1.7), **individual mutations**, like [S:E484K](https://outbreak.info/situation-reports?muts=S%3AE484K), or **groups of mutations**, like [S:E484K and S:P681R](https://outbreak.info/situation-reports?muts=S%3AE484K&muts=S%3AP681R). View the [Variant Tracker](varianttracker.html) and [Location Tracker](locationtracker.html) vignettes for more advanced examples.

---

### Cumulative variant prevalence geographically
To get the prevalence of a particular variant to compare between locations, you can access the data through the `getCumulativeBySubadmin` function. Note that there are options to supply a location, like `"United States"` to view the prevalence broken down by U.S. state, and/or over the past `n` days. See the [Variant Tracker Vignette](varianttracker.html) for more details.
```{r cumulative}
# Calculate cumulative prevalence of B.1.1.7 by country
b117_world = getCumulativeBySubadmin(pangolin_lineage = "B.1.1.7")

# filtering down the data to view a few countries
b117_world %>%
  filter(name %in% c("Canada", "United Kingdom", "Australia", "New Zealand")) %>%
  select(name, proportion, proportion_ci_lower, proportion_ci_upper) %>%
  arrange(desc(proportion)) %>%
  knitr::kable()
```

---

### Mutations within a lineage
When we say a lineage, like B.1.1.7, what does that actually mean?  The `getMutationsByLineage` function allows you to pull the prevalence of all the mutations within all the sequences assigned to B.1.1.7 or other lineages, and  `plotMutationHeatmap` allows you to compare their prevalence in a heatmap:
```{r characteristic_mutations}
char_muts = getMutationsByLineage(pangolin_lineage = c("B.1.1.7", "B.1.351", "B.1.617.2", "P.1"))
plotMutationHeatmap(char_muts, title = "Mutations with at least 75% prevalence in Variants of Concern", lightBorders = FALSE)
```

---

### Additional genomic functions
All the [Variant Prevalence Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-sars-cov-variant-prevalence) provide documentation on their functionality and examples.


---

## Cases & deaths data
Epidemiology data in outbreak.info includes daily cases and death data for World Bank Regions, individual countries, states/provinces, U.S. Metropolitan areas, and U.S. counties.

---


### Accessing Cases & Deaths data
All cases and death data can be accessed through the main function `getEpiData` and plotted with `plotEpiData`. For instance, you can compare the cases per capita in a few major metropolitan areas:
```{r epi}
# Get the epi data
epi_metro = getEpiData(name = c("Detroit-Warren-Dearborn, MI", "New Orleans-Metairie, LA"))

# Accessing just the first row:
t(epi_metro[1,])

# Get the epi data and plot it.
plotEpiData(locations = c("Detroit-Warren-Dearborn, MI", "New Orleans-Metairie, LA"), variable = "confirmed_rolling_per_100k")
```

Variables are described in the [Epidemiology Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/epidemiologyDataDictionary.html).

---

### Customizing epidemiology features
View the [Epidemiology vignette](outbreakvignette.html) for more details on how to access these data and [Epidemiology Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/epidemiologyDataDictionary.html) for information on the variables contained within the data.

### Additional epidemiology functions
All the [Cases & Deaths Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-cases-deaths) provide documentation on their functionality and examples.
