---
title: "Introduction to outbreakinfo"
# author: "Laura Hughes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to outbreakinfo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
[Outbreak.info](https://outbreak.info) is a project to enable the tracking of SARS-CoV-2 Variants within the COVID-19 pandemic. This R package offers access to the data we have gathered and calculated along with visualizations to replicate the visualizations on outbreak.info. Here, we'll outline some of the basic functionality of the R package to access the genomic (SARS-CoV-2 variant) data and epidemiology data (COVID-19 cases and deaths).

## SARS-CoV-2 variant prevalence

The core functionality within the suite of [Variant Prevalence Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-sars-cov-variant-prevalence) include accessing data for:

- [Variant prevalence over time](#prevalence-over-time)
- [Cumulative variant prevalence](#cumulative-variant-prevalence)
- [Calculating mutation prevalence within a variant](#muations-within-a-variant)

### Getting started
Before we start, to we need to provide our GISAID credentials to access the genomic data.
If you don't have a GISAID account, you can [register](https://www.gisaid.org/registration/register/)
for one on their website. It may take a day or two for the account to become active.
```{r pkgs, message = FALSE, results='hide'}
# Package imports
library(outbreakinfo)

# Not needed; just used to tidy some of the outputs
library(dplyr)
library(knitr) 
```

```{r auth, message = FALSE, results='hide', eval=FALSE}
# Authenticate yourself using your GISAID credentials.
authenticateUser()
```


### Prevalence over time
One important aspect of a variant to track is how its prevalence is changing over time. We can calculate this prevalence for any lineage, mutation, combination of mutations, or lineage with added mutations. We'll access the data and then plot the results.

#### Lineage over time: B.1.1.7
```{r prev-time}
# Function to grab all the data for the prevalence of B.1.1.7 in San Diego
b117_san_diego = getPrevalence(pangolin_lineage = "B.1.1.7", location="San Diego")

# Accessing just the first row:
t(b117_san_diego[1,])

# Plotting it:
plotPrevalenceOverTime(b117_san_diego, title = "B.1.1.7 prevalence in San Diego")
```
Variables are described in the [Genomics Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/genomicsDataDictionary.html).

#### Customizing variant inputs
In addition to viewing the change in prevalence of a lineage over time, more complex variants can be queried. This includes **combinations of lineages**, like the [Delta variant](https://outbreak.info/situation-reports/delta), **lineages with additional mutation(s)**, like [B.1.1.7 with S:E484K](https://outbreak.info/situation-reports?muts=S%3AE484K&pango=B.1.1.7), **individual mutations**, like [S:E484K](https://outbreak.info/situation-reports?muts=S%3AE484K), or **groups of mutations**, like [S:E484K and S:P681R](https://outbreak.info/situation-reports?muts=S%3AE484K&muts=S%3AP681R). View the [Variant Tracker](varianttracker.html) and [Location Tracker](locationtracker.html) vignettes for more advanced examples.

### Cumulative variant prevalence geographically
To get the prevalence of a particular variant to compare between locations, you can access the data through the `getCumulativeBySubadmin` function. Note that there are options to supply a location, like `"United States"` to view the prevalence broken down by U.S. state, and/or over the past `n` days. See the [Variant Tracker Vignette](varianttracker.html) for more details.
```{r cumulative}
# Calculate cumulative prevalence of B.1.1.7 by country
b117_world = getCumulativeBySubadmin(pangolin_lineage = "B.1.1.7")

# filtering down the data to view a few countries
b117_world %>% 
  filter(name %in% c("Canada", "United Kingdom", "Australia", "New Zealand")) %>%
  select(name, proportion, proportion_ci_lower, proportion_ci_upper) %>% 
  arrange(desc(proportion)) %>% 
  knitr::kable()
```


### Mutations within a variant

### Additional genomic functions
All the [Variant Prevalence Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-sars-cov-variant-prevalence) provide documentation on their functionality and examples.

## Cases & deaths data
Epidemiology data in outbreak.info includes daily cases and death data for World Bank Regions, individual countries, states/provinces, U.S. Metropolitan areas, and U.S. counties.

### Accessing Cases & Deaths data
All cases and death data can be accessed through the main function `getEpiData` and plotted with `plotEpiData`. For instance, you can compare the cases per capita in a few major metropolitan areas:
```{r epi}
# Get the epi data
epi_metro = getEpiData(name = c("Detroit-Warren-Dearborn, MI", "New Orleans-Metairie, LA"))

# Accessing just the first row:
t(epi_metro[1,])

# Get the epi data and plot it.
plotEpiData(locations = c("Detroit-Warren-Dearborn, MI", "New Orleans-Metairie, LA"), variable = "confirmed_rolling_per_100k")
```

Variables are described in the [Epidemiology Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/epidemiologyDataDictionary.html).

### Customizing epidemiology features
View the [Epidemiology vignette](outbreakvignette.html) for more details on how to access these data and [Epidemiology Data Dictionary](https://outbreak-info.github.io/R-outbreak-info/reference/epidemiologyDataDictionary.html) for information on the variables contained within the data.

### Additional epidemiology functions
All the [Cases & Deaths Functions](https://outbreak-info.github.io/R-outbreak-info/reference/index.html#section-cases-deaths) provide documentation on their functionality and examples.
