---
title: "outbreak.info Variant Tracker"
# author: "Laura Hughes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{outbreak.info Variant Tracker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Accessing the data in outbreak.info's Variant Tracker
Outbreak.info's Variant Tracker allows you to view how combinations of lineages and mutations are changing over time. These reports allow you to answer questions like:

- [What mutations are present within that lineage?](#what-mutations-are-characteristic-of-the-variant-)
- [How prevalent is a variant overall?](#how-prevalent-is-a-variant-overall-)
- [How is the prevalence of the variant changing?](#how-is-the-prevalence-of-a-variant-changing-)
- [Where is the variant most prevalent?](#where-is-the-variant-most-prevalent-)
- [What is known about the variant in the research literature?](#what-do-we-know-about-the-variant-in-the-literature-)


Here, we'll demonstrate how to access the data in the [P.1 Lineage Report](https://outbreak.info/situation-reports?pango=P.1&loc=BRA&loc=USA&selected=Worldwide&overlay=false). The P.1 / Gamma lineage was labeled a [Variant of Concern](https://outbreak.info/situation-reports) by the WHO in January 2021 because it showed evidence of increased transmissibility, virulence, and/or decreased diagnostic, therapeutic, or vaccine efficacy.

## Before we start...
Import the packages we'll use and provide our GISAID credentials to access the data.
If you don't have a GISAID account, you can [register](https://www.gisaid.org/registration/register/)
for one on their website. It may take a day or two for the account to become active.
```{r pkgs, message = FALSE, results='hide'}
# Package imports
library(dplyr)
library(purrr)
library(ggplot2)
library(knitr)
library(outbreakinfo)
```

```{r auth, message = FALSE, results='hide', eval=FALSE}
# Authenticate yourself using your GISAID credentials.
authenticateUser()
```


## What mutations are characteristic of the variant?
A very basic question: what is P.1? What mutations consistently appear in most sequences within the lineage?

### Grabbing the characteristic mutations
```{r characteristic_mutations}
p1 = getMutationsByLineage(pangolin_lineage="P.1")
plotMutationHeatmap(p1, title="Characteristic mutations of P.1 occurring in at least 75% of sequences")
```

### Comparing to other lineages
We can also compare how the mutations of P.1 compare to its sublineages, to look how any mutations may have been aquired or disappeared. We can also lower (or raise) the threshold for which mutations are included
```{r all_gamma_mutations}
# First: find which sublineages are associated with P.1:
gamma_lineages = lookupSublineages("Gamma", returnQueryString = FALSE)

# Collect the data
gamma_mutations = getMutationsByLineage(pangolin_lineage=gamma_lineages, frequency=0.5, logInfo = FALSE)

# Plot!
plotMutationHeatmap(gamma_mutations, title = "S-gene mutations in > 50% of Gamma sequences")

# You can also change the gene you select by changing 
# plotMutationHeatmap(gamma_mutations, title = "ORF1a-gene mutations in Gamma lineages", gene2Plot="ORF1a")
```


## How prevalent is a variant overall?
To start with, let's get some basic information about the overall, cumulative prevalence of P.1 in certain locations. We'll look at the worldwide totals, in Brazil, in the United States, and in Illinois.
```{r cum_prevalence}
locations = c("Brazil", "United States", "Illinois")

worldwide_prevalence = getPrevalence(pangolin_lineage = "P.1", cumulative = TRUE)
cumulative_prevalences = purrr::map_df(locations, function(loc) getPrevalence(pangolin_lineage = "P.1", location = loc, cumulative = TRUE)) %>% 
  arrange(desc(value.global_prevalence))

cumulative_prevalences = worldwide_prevalence %>% dplyr::bind_rows(cumulative_prevalences)

kable(cumulative_prevalences %>% dplyr::select(location, value.global_prevalence, value.lineage_count, value.first_detected, value.last_detected))
```


## How is the prevalence of a variant changing?

## Where is the variant most prevalent?


## What do we know about the variant in the literature?
The pace of research on COVID-19 has been staggering, with the research community publishing thousands of papers, clinical trials, datasets and more each week. The [P.1 papers and other resources](https://outbreak.info/resources/search?q=%22P.1%22) can be found in outbreak.info's Research Library. At this time, the R package does not provide access to these data, but they can be found in our [resource API](https://api.outbreak.info/try/resources).
